preciseTAD: A machine learning framework for precise TAD boundary prediction at base level resolution

### 2020-08-23 Manuscript/manuscript.Rmd
edited figure legends and numbering

Next, we evaluated the biological significance of *preciseTAD* boundaries. Peak regions associated with known molecular drivers of 3D chromatin, including CTCF, RAD21, SMC3, and ZNF143, were found to co-localize more frequently around *preciseTAD* predicted boundaries (Figure 5B). Furthermore, signal heatmaps indicated that our predicted boundaries were more enriched for the same genomic annotations than either Arrowhead or Peakachu boundaries alone (Figure 5 C-D; Supplementary Figures 5-6). When compared to called TAD or loop boundaries, *preciseTAD* boundaries were statistically significantly closer to the same cell line-specific top-ranked TFBS (p<0.001 and p<0.001 versus Arrowhead and Peakachu respectively) (Figure 5E, Supplementary Figure 7). These results indicate that *preciseTAD*-predicted boundaries better reflected the known biology of boundary formation.

@@

TAD boundaries are relatively conserved across cell lines [@dixon2012topological; @krefting2018evolutionary; @nora2012spatial; @sexton2012three]. To assess conservation, we evaluated the overlap between cell line-specific boundaries detected by *preciseTAD*, Arrowhead, and Peakachu after flanking them by the resolution of the Hi-C data. Boundaries involving Arrowhead were flanked by 5 kb, while boundaries involving Peakachu were flanked by 10 kb. We found that 35% of *preciseTAD* predicted TAD boundaries and 55% of *preciseTAD* predicted loop boundaries were conserved between cell lines, compared to only 26% and 55% for Arrowhead and Peakachu alone (Figure 5F). The better conservation of *preciseTAD*-predicted boundaries further supports the notion of their higher biological relevance.

![**preciseTAD boundaries are enriched for known molecular drivers of 3D chromatin.** (A) The location of preciseTAD (green) vs. Arrowhead (blue) boundaries on GM12878 (chr 14: 50085000-50800000). The black line represents the predicted probability of each base being a TAD boundary. (B) Genome browse visualization of bigwig files for known molecular drivers of 3D chromatin including CTCF, RAD21, SMC3, and ZNF143. (C) Enriched heatmaps comparing the enrichment of CTCF signal around Arrowhead TAD boundaries and preciseTAD boundaries, and (D) around Peakachu loop boundaries and preciseTAD for the entire genome. (E) Log 2 genomic distances from boundaries to the nearest CTCF sites. (F) Venn diagrams illustrating the levels of conservation between TAD/loop boundaries for GM12878 (red) and K562 (blue) identified by preciseTAD (using Arrowhead; top left), Arrowhead (top right), preciseTAD (using Peakachu; bottom left), and Peakachu (bottom right). The median Jaccard Index (IQR) across each chromosome is provided below each Venn diagram](Figures/figure5.png)

@@

The majority of boundaries predicted by the Arrowhead-trained *preciseTAD* model represented a subset of the larger group of boundaries predicted by the Peakachu-trained model. We found that 95.6% and 95.7% of our predicted TAD boundaries were overlapped by predicted loop boundaries for GM12878 and K562, respectively (Supplementary Figure 8). This is expected as loop boundaries (Peakachu) are more abundant, while TAD boundaries (Arrowhead) represent the narrower level of the 3D chromatin organization. Together with the results in Figure 4, the high overlap between boundaries predicted by Arrowhead- and Peakachu-trained models suggests that TAD and loop domain boundaries may be driven by similar molecular mechanisms.

@@

*preciseTAD* offers flexibility in controlling both the number of predicted boundaries and the distance between them. The two primary parameters are the probability threshold $t$ and $\epsilon$ (referred to as eps; parameter of DBSCAN). The combination of these two quantities changes the resulting number of predicted boundaries from *preciseTAD*. Lower values of $t$ and $\epsilon$ will result in more clusters of bases, and therefore, more boundaries. As a heuristic, we evaluated the pairwise combination of four different thresholds ($t=(0.95,0.975,0.99,1.0)$) and four different $eps$ values ($\epsilon=(r, 2r, 3r, 4r, 5r)$)), where $r=5000$, the optimal resolution. We found that the normalized overlaps--calculated as the total number of ChIP-seq peaks that overlapped within a given flanked boundary, divided by the number of boundaries --between top TFBS sites and flanked preciseTAD boundaries were maximized for combinations of $t=1.0$ and $eps=r$ (Supplementary Figure 9).

@@

**Supplementary Figure 5. preciseTAD boundaries are more enriched for known molecular drivers of 3D chromatin.** Enriched heatmaps comparing the enrichment of preciseTAD (green) boundaries with Arrowhead called TAD boundaries (blue) around key (A) GM12878-specific and (B) K562-specific transcription factor binding sites known to contribute to the 3D chromatin architecture.

**Supplementary Figure 6. preciseTAD boundaries are more enriched for known molecular drivers of 3D chromatin.** Enriched heatmaps comparing the enrichment of preciseTAD (green) boundaries with Peakachu predicted chromatin loop boundaries (blue) around key (A) GM12878-specific and (B) K562-specific transcription factor binding sites known to contribute to the 3D chromatin architecture.

**Supplementary Figure 7. preciseTAD boundaries are spatially closer to known molecular drivers of 3D chromatin.** Boxplots comparing the $log_{2}$ genomic distances from predicted and called boundaries to the nearest (A) GM12878-specific and (B) K562-specific CTCF, RAD21, SMC3, and ZNF143 transcription factor binding sites.

**Supplementary Figure 8. preciseTAD predicted TAD boundaries using Arrowhead are nested within preciseTAD predicted chromatin loop boundaries using Peakachu.** Venn diagrams illustrating the agreement between boundaries predicted using Arrowhead- and Peakachu-trained models. TAD/loop boundaries were flanked by resolution (5 kb for models using Arrowhead and 10 kb for Peakachu).

**Supplementary Figure 9.** Linecharts illustrating the normalized overlap between known molecular drivers of 3D chromatin and flanked predicted TAD boundaries identified by preciseTAD (using Arrowhead ground truth boundaries) for different combinations of thresholds (t) and epsilon parameter values (eps) on GM12878 and for chromosome 22. Normalized overlap was calculated as the total number of ChIP-seq peaks that overlapped within a given flanked boundary, divided by the number of boundaries that were predicted (on chr22). Boundaries were flanked by 5 kb resolution.

>>>>

Next, we evaluated the biological significance of *preciseTAD* boundaries.

Peak regions associated with known molecular drivers of 3D chromatin, including CTCF, RAD21, SMC3, and ZNF143, were found to co-localize more frequently around *preciseTAD* predicted TAD and loop boundaries compared to Arrowhead called TAD and Peakached loop boundaries respectively (Figure 5B-5C; Supplementary Figure 5). Furthermore, signal heatmaps indicated that our predicted boundaries were more enriched for the same genomic annotations than either Arrowhead or Peakachu boundaries alone (Supplementary Figures 6-7). When compared to called TAD or loop boundaries, *preciseTAD* boundaries were statistically significantly closer to the same cell line-specific top-ranked TFBS (p<0.001 and p<0.001 versus Arrowhead and Peakachu respectively) (Figure 6A, Supplementary Figure 8). These results indicate that *preciseTAD*-predicted boundaries better reflected the known biology of boundary formation.

![**preciseTAD boundaries are enriched for known molecular drivers of 3D chromatin.** (A) The location of ARROWHEAD called TAD boundaries (blue) vs *preciseTAD* predicted TAD boundaries (green) on GM12878 (chr 14: 50085000-50800000). The black line represents the predicted probability of each base being a TAD boundary. Genome browser visualization of bigwig files for known molecular drivers of 3D chromatin including CTCF, RAD21, SMC3, and ZNF143 is presented underneath. Profile plots comparing the signal strength of top TFBS around flanked (B) ARROWHEAD called TAD boundaries (blue) and *preciseTAD* predicted TAD boundaries (green), and (C) around Peakachu loop boundaries (red) and *preciseTAD* predicted loop boundaries (green).](Figures/figure5.png)

@@

TAD boundaries are relatively conserved across cell lines [@dixon2012topological; @krefting2018evolutionary; @nora2012spatial; @sexton2012three]. To assess conservation, we evaluated the overlap between cell line-specific boundaries detected by *preciseTAD*, Arrowhead, and Peakachu after flanking them by the resolution of the Hi-C data. Boundaries involving Arrowhead were flanked by 5 kb, while boundaries involving Peakachu were flanked by 10 kb. We found that 35% of *preciseTAD* predicted TAD boundaries and 55% of *preciseTAD* predicted loop boundaries were conserved between cell lines, compared to only 26% and 55% for Arrowhead and Peakachu alone (Figure 6B). The better conservation of *preciseTAD*-predicted boundaries further supports the notion of their higher biological relevance.

![**preciseTAD predicted boundaries are both statistically significantly closer to CTCF sites and more conserved across cell lines**. (A) Log2 genomic distances from called and predicted boundaries to nearest CTCF sites. The p-value is derived from the Wilcoxon Rank Sum test. (B) Venn diagrams illustrating the levels of conservation between TAD/loop boundaries for GM12878 (red) and K562 (blue) identified by *preciseTAD* (using ARROWHEAD; top left), ARROWHEAD (top right), *preciseTAD* (using PEAKACHU; bottom left), and PEAKACHU (bottom right). The median Jaccard Index (IQR) across each chromosome is provided below each venn diagram.](Figures/figure6.png)

@@

The majority of boundaries predicted by the Arrowhead-trained *preciseTAD* model represented a subset of the larger group of boundaries predicted by the Peakachu-trained model. We found that 95.6% and 95.7% of our predicted TAD boundaries were overlapped by predicted loop boundaries for GM12878 and K562, respectively (Supplementary Figure 9). This is expected as loop boundaries (Peakachu) are more abundant, while TAD boundaries (Arrowhead) represent the narrower level of the 3D chromatin organization. Together with the results in Figure 4, the high overlap between boundaries predicted by Arrowhead- and Peakachu-trained models suggests that TAD and loop domain boundaries may be driven by similar molecular mechanisms.

@@

*preciseTAD* offers flexibility in controlling both the number of predicted boundaries and the distance between them. The two primary parameters are the probability threshold $t$ and $\epsilon$ (referred to as eps; parameter of DBSCAN). The combination of these two quantities changes the resulting number of predicted boundaries from *preciseTAD*. Lower values of $t$ and $\epsilon$ will result in more clusters of bases, and therefore, more boundaries. As a heuristic, we evaluated the pairwise combination of four different thresholds ($t=(0.95,0.975,0.99,1.0)$) and four different $eps$ values ($\epsilon=(r, 2r, 3r, 4r, 5r)$)), where $r=5000$, the optimal resolution. We found that the normalized overlaps--calculated as the total number of ChIP-seq peaks that overlapped within a given flanked boundary, divided by the number of boundaries --between top TFBS sites and flanked *preciseTAD* boundaries were maximized for combinations of $t=1.0$ and $eps=r$ (Supplementary Figure 10).

@@

**Supplementary Figure 5.** Profile plots comparing the signal strength of top TFBS around flanked (B) ARROWHEAD called TAD boundaries (blue) and *preciseTAD* predicted TAD boundaries (green), and (C) around Peakachu loop boundaries (red) and preciseTAD predicted loop boundaries (green) on K562.

**Supplementary Figure 6.** *preciseTAD* boundaries are more enriched for known molecular drivers of 3D chromatin. Enriched heatmaps comparing the enrichment of (A) ARROWHEAD called TAD boundaries vs *preciseTAD* predicted TAD boundaries, and (B) Peakachu loop boundaries vs *preciseTAD* predicted loop boundaries around key GM12878-specific transcription factor binding sites known to contribute to the 3D chromatin architecture.

**Supplementary Figure 7.** *preciseTAD* boundaries are more enriched for known molecular drivers of 3D chromatin. Enriched heatmaps comparing the enrichment of (A) ARROWHEAD called TAD boundaries vs *preciseTAD* predicted TAD boundaries, and (B) Peakachu loop boundaries vs preciseTAD predicted loop boundaries around key K562-specific transcription factor binding sites known to contribute to the 3D chromatin architecture.

**Supplementary Figure 8.** *preciseTAD* boundaries are spatially closer to known molecular drivers of 3D chromatin. Boxplots comparing the log2 genomic distances from predicted and called boundaries to the nearest (A) GM12878-specific and (B) K562-specific CTCF, RAD21, SMC3, and ZNF143 transcription factor binding sites. The p-value is derived from the Wilcoxon Rank Sum test.

**Supplementary Figure 9.** *preciseTAD* predicted TAD boundaries using ARROWHEAD are nested within *preciseTAD* predicted chromatin loop boundaries using PEAKACHU. Venn diagrams illustrating the agreement between boundaries predicted using ARROWHEAD- and Peakachu-trained models. TAD/loop boundaries were flanked by resolution (5 kb for models using ARROWHEAD and 10 kb for PEAKACHU).

**Supplementary Figure 10.** Linecharts illustrating the normalized overlap between known molecular drivers of 3D chromatin and flanked predicted TAD boundaries identified by *preciseTAD* (using Arrowhead ground truth boundaries) for different combinations of thresholds (t) and epsilon parameter values (eps) on GM12878 and for chromosome 22. Normalized overlap was calculated as the total number of ChIP-seq peaks that overlapped within a given flanked boundary, divided by the number of boundaries that were predicted (on chr22). Boundaries were flanked by 5 kb resolution.
